---
title: "Activity Monitoring Data Project"
author: "Cathy Staats"
date: "March 9, 2017"
output: 
  html_document:
    keep_md: TRUE

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

#### Read in activity monitoring dataset and produce print a summary

```{r importdata}
activity <- read.csv("activity.csv")
summary(activity)
```

#### Summarize steps by date

```{r sumbydate}
library(dplyr)
StepsbyDate <- select(activity, date, steps) %>%
             group_by(date) %>%
             summarize(steps = sum(steps, na.rm=TRUE))
dim(StepsbyDate)
```

### Plot histogram of number of days by steps

```{r plothist, fig.keep='high', fig.path='figure/'}
library(ggplot2)
ggplot(StepsbyDate, aes(steps)) + 
        geom_histogram(binwidth = 1000) +
        scale_x_continuous("Total Daily Steps") +
        scale_y_continuous("Number of Days") +
        ggtitle("Days by Number of Steps") 
```

### Display mean and median of steps taken by day
```{r statsbyday}
mean(StepsbyDate$steps, na.rm = TRUE)
median(StepsbyDate$steps, na.rm = TRUE)
```

#### Summarize data to get average steps by 5-minute interval
```{r SumbyInterval}
library(dplyr)
StepsbyInt <- select(activity, interval, steps) %>%
             group_by(interval) %>%
             summarize(steps = mean(steps, na.rm = TRUE))

```
### Plot average steps taken by 5-minute interval averaged across days

```{r plotbyint, fig.keep='high', fig.path='figure/'}

library(ggplot2)
ggplot(StepsbyInt, aes(interval, steps)) + 
     geom_line() +
     scale_x_continuous("5-minute Interval of Day") +
     scale_y_continuous("Average Steps") +
     ggtitle("Steps by 5-Minute Interval") 
```

### Determine which interval has the max number of steps

```{r findmax}

maxint <- which.max(StepsbyInt$steps)
StepsbyInt$interval[maxint]
```

#### Determine number of missing values in data

```{r findmissing}
sum(is.na(activity$steps))
```

### Inpute missing data
Split dataframe into two: rows with missing data and
rows without missing data. Replace missing data with average based
on the average for that 5-minute interval across all of the days. Using sqldf to join based on interval.

```{r splitframe}
library(dplyr)
NArows <- filter(activity, is.na(steps))
dim(NArows)

NotNArows <- filter(activity, !is.na(steps))
dim(NotNArows)
```

```{r fillmissing}
StepsbyInt <- transform(StepsbyInt, 
                steps = as.integer(round(steps, digits = 0)))
library(sqldf)

join_string <- "select
                StepsbyInt.steps,
                NArows.date,
                NArows.interval
from NArows
left join StepsbyInt
on NArows.interval = StepsbyInt.interval"

NAfilled <- sqldf(join_string,stringsAsFactors = FALSE)

## Combine with rows that did not having missing values 
activitynm <- as.data.frame(rbind(NAfilled, NotNArows))
dim(activitynm)
## double check that there are not no missing values
sum(is.na(activitynm$steps))
```

#### Summarize steps by date using data after imputing missing values

```{r sumbydatenm}
library(dplyr)
StepsbyDate2 <- select(activitynm, date, steps) %>%
             group_by(date) %>%
             summarize(steps = sum(steps))
dim(StepsbyDate2)
```

### Plot histogram of number of days by steps using data after imputing missing values

```{r plothistnm, fig.keep='high', fig.path='figure/'}
library(ggplot2)
ggplot(StepsbyDate2, aes(steps)) + 
        geom_histogram(binwidth = 1000) +
        scale_x_continuous("Total Daily Steps") +
        scale_y_continuous("Number of Days") +
        ggtitle("Days by Number of Steps") 
```

#### Display mean and median of steps taken by day before imputation
  
```{r statsbydaynm}
mean(StepsbyDate$steps, na.rm = TRUE)
median(StepsbyDate$steps, na.rm = TRUE)
```

#### Display mean and median of steps taken by day after imputation
```{r statsbydaynm2}
mean(StepsbyDate2$steps)
median(StepsbyDate2$steps)
```

#### Display differnce in mean and median of steps taken by day
####  after imputing missing values minus before
```{r statsbydaynm3}
mean(StepsbyDate2$steps) - mean(StepsbyDate$steps, na.rm = TRUE)
median(StepsbyDate2$steps) - median(StepsbyDate$steps, na.rm = TRUE)

```
##### Mean number of steps increased by about 1400 steps while median increased by 367 steps.

#### Add weekday / weekend as a factor variable to the dataset
```{r weekdayftr}
library(dplyr)

activitynm <- transform(activitynm, date = as.Date(date))

activitynm <- 
  mutate(activitynm, dt = weekdays(date))

Weekends <- 
   filter(activitynm, dt == 'Sunday' | dt == 'Saturday') %>%
   mutate (dt = 'Weekend')

Weekdays <- 
   filter(activitynm, dt != 'Sunday' & dt != 'Saturday') %>%
   mutate (dt = 'Weekday')

allweek <- rbind(Weekends, Weekdays)
```

#### Summarize data to get average steps by 5-minute interval
   by Weekend vs weekday
```{r SumbyIntervalbw}
library(dplyr)
StepsbyInt <- select(allweek, dt, interval, steps) %>%
             group_by(dt, interval) %>%
             summarize(steps = mean(steps))
```

### Plot average steps taken by 5-minute interval averages across days
#### by weekend vs. weekday
```{r plotbydaytype, fig.keep='high', fig.path='figure/'}
library(ggplot2)
ggplot(StepsbyInt, aes(interval, steps)) + 
     geom_line() +
     facet_wrap(~dt, ncol=1) +
     scale_x_continuous("5-minute Interval of Day") +
     scale_y_continuous("Average Steps") +
     ggtitle("Steps by 5-Minute Interval") 
```


